---
title: "Cognitive Warrant Analytic Pipeline"
subtitle: "Walkthrough of the proof-of-concept analysis, for transparency and reproducibility"
---

## Overview

This page is a narrated version of the analysis pipeline described in Chapter 5 of my [research document](manuscript.qmd) ("The Analytic Pipeline: Logistic Regression and the Estimation of Cognitive Warrant Metrics"). It runs the pipeline on the real merged Experiment 2 / LCSH dataset and is published here, in the style of Trott and Bergen's own [pre-registration code](https://github.com/seantrott/trott_ph_amb), so that other researchers can inspect, reproduce, and critique every step -- including the diagnostic dead ends and corrections that shaped the final approach.

The pipeline takes each **stimulus pair** (a prime sentence and a target sentence sharing a word, drawn from Trott and Bergen's Experiment 2) and asks how much information a predictor -- BERT cosine distance ($d$), LCSH heading congruence ($S$), or LCSH hierarchical distance ($h$) -- carries about whether the pair reflects the *same* underlying sense ($Y$). That question is operationalized as **cognitive information**, $CI(d)$: the reduction in the entropy of $Y$ once the predictor is known, expressed in bits.

At a high level, the pipeline:

1. Restructures trial-level records (repeated measures) into one row per unique stimulus pair.
2. Runs diagnostics that determine how the rest of the pipeline should be interpreted -- most importantly, whether the LCSH-derived predictor $S$ is empirically informative or a tautological restatement of the outcome.
3. Fits baseline (`Y ~ d`) and full (`Y ~ d + S`) logistic models, using Firth's bias-reduced estimator where ordinary `glm()` shows separation.
4. Converts fitted probabilities to entropy, then to piecewise and pointwise cognitive information.
5. Repeats the same pipeline with `hier_distance` (recoded as `h_finite`) in place of `S`, since `S` turns out to be tautologically identical to the outcome in this dataset (see [Data diagnostics]).
6. Saves all tables, plots, and fitted models used elsewhere in the write-up.

Required packages: `install.packages(c("tidyverse", "logistf", "lme4", "splines", "patchwork", "knitr"))`.

```{r setup}
#| message: false

library(tidyverse) # data wrangling and plotting
library(logistf)   # Firth's bias-reduced logistic regression (stable under separation)
library(lme4)      # generalized linear mixed-effects models (random-intercept robustness check)
library(splines)   # loaded for parity with the modeling scripts that build on this one
library(patchwork) # combines separate ggplots into one figure
library(knitr)      # renders the LaTeX table at the end of this page

set.seed(1226)
```

## Step 0: Load and restructure trial-level records

The trial-level file inherits the repeated-measures structure of Trott and Bergen's Experiment 2: 187 subjects, each responding to a subset of stimulus pairs. This step collapses that structure down to one row per unique `(word, version_with_order)` pair, which is the unit of analysis for everything that follows.

This walkthrough reads `tb2023_exp2_lcsh_distances.csv` from `assets/data/`, the copy of the merged dataset committed to the public repository (the working copy under `data/processed/`, produced by [`get_distances.py`](https://github.com/nicholasmeister/cognitive-warrant/blob/main/src/analysis/get_distances.py), is regenerated locally and excluded from version control). Paths below are relative to this file's location in `pages/`, so the page renders correctly both locally and as part of the built site.

```{r load-data}
#| message: false

raw <- read_csv("../assets/data/tb2023_exp2_lcsh_distances.csv", show_col_types = FALSE)

# distance_elmo_lcsh is entirely missing in this file (the LCSH-anchored ELMo
# frame was never computed) -- dropped rather than carried as an all-NA column.
raw <- raw %>% select(-distance_elmo_lcsh)
```

`same`, `same_heading`, `hier_distance`, `distance_bert`, `ambiguity_type`, and `Class` are all properties of the *stimulus pair*, not of the trial -- each should take exactly one value within every `(word, version_with_order)` group. The check below confirms that empirically rather than assuming it; if it ever fails on a future data pull, that is a signal to stop and investigate before trusting the collapse that follows.

```{r pair-level-check}
pair_key <- c("word", "version_with_order")
pair_level_cols <- c("same", "same_heading", "hier_distance", "same_nbhd",
                      "distance_bert", "distance_bert_lcsh", "distance_elmo",
                      "ambiguity_type", "Class")

nun_check <- raw %>%
  group_by(across(all_of(pair_key))) %>%
  summarise(across(all_of(pair_level_cols), n_distinct), .groups = "drop")

stopifnot(all(nun_check[pair_level_cols] == 1))
message("Pair-level covariates confirmed constant within (word, version_with_order).")
```

With that confirmed, each pair is reduced to a single row. A trial-level summary (mean accuracy, mean log RT, subject/trial counts) is kept alongside it in case RT or accuracy covariates are wanted for a robustness check later -- it is not used by the $CI(d)$ pipeline itself, which operates on $Y_i$ = `same`, not on RT or accuracy.

```{r restructure}
trial_summary <- raw %>%
  group_by(across(all_of(pair_key))) %>%
  summarise(
    n_trials     = n(),
    n_subjects   = n_distinct(subject),
    mean_correct = mean(correct_response),
    mean_log_rt  = mean(log_rt),
    .groups = "drop"
  )

pairs_df <- raw %>%
  distinct(across(all_of(c(pair_key, pair_level_cols)))) %>%
  left_join(trial_summary, by = pair_key) %>%
  mutate(
    pair_id = paste(word, version_with_order, sep = "__"),
    Y       = as.integer(same),          # Y_i: ground-truth same-sense indicator
    S       = as.integer(same_heading),  # S_i: LCSH heading congruence
    d       = distance_bert,             # d_i: BERT cosine distance (unmodified frame)
    h       = hier_distance              # h_i: BT/NT shortest-path distance
                                          #      (Inf if S_i = 0 with no shared ancestor
                                          #      within the search horizon)
  )

message(sprintf(
  "Collapsed %d trial-level rows (%d subjects) into %d unique stimulus pairs across %d words.",
  nrow(raw), n_distinct(raw$subject), nrow(pairs_df), n_distinct(pairs_df$word)
))
```

## Data diagnostics

These checks run *before* any downstream number is trusted, because they determine how the rest of the pipeline should be read.

### Is $S$ perfectly collinear with $Y$?

This is the single most consequential check in the pipeline. If `same_heading` ($S$) is identical to `same` ($Y$) for every pair, the "full model" logistic regression of $Y$ on $d$ and $S$ is not testing whether LCSH heading congruence adds information beyond BERT distance -- it is recovering a tautology, because $S$ was constructed to equal $Y$ for every pair in this file (each dictionary sense M1/M2 was mapped one-to-one onto a single LCSH authority record). Under that condition, $CI(d)$ comes out equal to $H(Y \mid d)$ exactly (up to numerical floor), and warrant efficiency $\bar\eta$ is $\approx 1$ everywhere -- a mechanical consequence of how `same_heading` was derived here, not an empirical finding about LCSH's cognitive value.

```{r diagnostic-s-vs-y}
same_vs_heading <- mean(pairs_df$Y == pairs_df$S)
message(sprintf("Proportion of pairs where Y (same) == S (same_heading): %.4f", same_vs_heading))

if (isTRUE(all.equal(same_vs_heading, 1))) {
  warning(paste(
    "same_heading is IDENTICAL to same for every pair in this file.",
    "The baseline-vs-full model comparison below will show a trivial,",
    "perfectly-separated effect of S -- this reflects how the LCSH mapping",
    "was constructed (one heading per dictionary sense), not an empirical",
    "test of whether LCSH boundaries carry information beyond distance.",
    "hier_distance (h), which DOES vary independently of Y among",
    "cross-heading pairs, is the more informative structural variable for",
    "this proof of concept -- see 'Structural extension' below."
  ))
}
```

Confirmed: `same_heading` and `same` agree on every pair in this file. Everything reported under "S (same_heading)" below should be read as a **methods validation** (does the pipeline recover the expected ceiling case correctly?), not as an empirical test of LCSH's cognitive value. The empirical test is the `hier_distance` extension further down.

### Where does `hier_distance` carry information?

`hier_distance` is `Inf` for cross-heading pairs with no shared ancestor within the search horizon, and exactly `0` for same-heading pairs (trivially, since a node has zero path length to itself). It is only *non-trivially* informative within the cross-heading ($S = 0$) subset, where it takes finite values greater than 0 for pairs that share a broader term, and `Inf` for pairs that do not.

```{r diagnostic-hier-distance}
message("hier_distance cross-tab (S = same_heading x finite/inf/zero):")
pairs_df %>%
  mutate(h_cat = case_when(
    is.infinite(h) ~ "inf",
    h == 0         ~ "zero",
    TRUE           ~ "finite>0"
  )) %>%
  count(S, h_cat)
```

### Coverage of the theoretically critical distance range

Chapter 4 (H3-as-theorem) predicts $CI(d)$ should peak somewhere in an intermediate zone of BERT cosine distance (provisionally, $d \in [0.30, 0.60]$). This just checks that the dataset has enough pairs in that range to say anything about it.

```{r diagnostic-distance-range}
message(sprintf("distance_bert range: [%.3f, %.3f]", min(pairs_df$d), max(pairs_df$d)))
message(sprintf(
  "Proportion of pairs with d in [0.30, 0.60]: %.3f",
  mean(pairs_df$d >= 0.30 & pairs_df$d <= 0.60)
))
```

## Step 1: Fit baseline and full logistic models

The baseline model estimates $P(Y=1 \mid d)$; the full model adds $S$:

$$\text{logit } P(Y=1\mid d) = \beta_0 + \beta_1 d \qquad\qquad \text{logit } P(Y=1\mid d, S) = \gamma_0 + \gamma_1 d + \gamma_2 S$$

Given the diagnostic above, the full model is expected to show quasi/perfect separation on $S$: `glm()` will warn that fitted probabilities are numerically 0 or 1, and the standard error on $\gamma_2$ will be enormous. The point estimates and fitted probabilities are still usable for the entropy calculations that follow, but the reported p-values and confidence intervals for $\gamma_2$ are not trustworthy from `glm()` alone in this degenerate case.

```{r fit-glm-models}
#| warning: true

m_baseline <- glm(Y ~ d, data = pairs_df, family = binomial(link = "logit"))
m_full_glm <- glm(Y ~ d + S, data = pairs_df, family = binomial(link = "logit"))

if (!m_full_glm$converged || any(abs(coef(m_full_glm)) > 15)) {
  message("m_full_glm shows signs of (quasi-)separation, as expected -- fitting Firth's bias-reduced logistic regression as a stable alternative.")
}
```

Firth's bias-reduced logistic regression does not blow up under perfect separation and yields finite, interpretable fitted probabilities even when $S$ is a near-deterministic function of $Y$. It is the preferred specification for the entropy calculations in the rest of this page whenever `m_full_glm` shows separation, as it does here.

```{r fit-firth-model}
m_full_firth <- logistf(Y ~ d + S, data = pairs_df)

summary(m_baseline)
summary(m_full_glm)
summary(m_full_firth)
```

A likelihood-ratio test gives the regression-side estimate of whether $CI(d) > 0$, i.e., the finite-sample test of $I(Y; S \mid d) = 0$. Reported for completeness; interpret cautiously given the separation diagnosed above.

```{r lrt-baseline-vs-full}
anova(m_baseline, m_full_glm, test = "Chisq")
```

As an optional robustness check, a random intercept for `word` accounts for pairs being nested within word (25 words, ~12 pairs each). This is likely to be singular or unstable for the same separation reason, but is worth inspecting.

```{r glmer-robustness-check}
#| warning: true

m_full_glmer <- tryCatch(
  glmer(Y ~ d + S + (1 | word), data = pairs_df, family = binomial),
  warning = function(w) { message("glmer warning (expected under separation): ", conditionMessage(w)); NULL },
  error   = function(e) { message("glmer error: ", conditionMessage(e)); NULL }
)
```

Finally, fitted probabilities are attached to `pairs_df`. The Firth model supplies the full-model probability as the primary specification; the `glm()` fit is retained alongside for comparison.

```{r fitted-probabilities}
pairs_df <- pairs_df %>%
  mutate(
    p_baseline = predict(m_baseline, type = "response"),
    p_full_glm = predict(m_full_glm, type = "response"),
    p_full     = plogis(predict(m_full_firth, newdata = pairs_df, type = "link"))
  )
```

## Step 2: From fitted probabilities to conditional entropy

The binary entropy function $H(p) = -p\log_2 p - (1-p)\log_2(1-p)$ converts a fitted probability into bits of uncertainty about $Y$.

```{r entropy-function}
h_b <- function(p) {
  p <- pmin(pmax(p, 1e-12), 1 - 1e-12)  # guard against log(0)
  -p * log2(p) - (1 - p) * log2(1 - p)
}

pairs_df <- pairs_df %>%
  mutate(
    H_baseline = h_b(p_baseline),
    H_full     = h_b(p_full)
  )
```

## Step 3: Piecewise (binned) cognitive information, $CI(d)$

Pairs are grouped into `K = 6` bins of BERT cosine distance -- fewer than in the schematic Chapter 5 template, given only 299 pairs in this dataset; re-run with `K = 8` or `K = 10` as a sensitivity check if extending this analysis.

```{r piecewise-ci}
K <- 6
pairs_df <- pairs_df %>%
  mutate(distance_bin = cut(d, breaks = K, include.lowest = TRUE))
```

Within each bin, `CI_k` is computed two ways. `H_bar_d` (and the resulting `CI_k`) comes from `m_baseline`, a model fit *independently* of `m_full_firth`, so this version is not guaranteed non-negative by Jensen's inequality -- that guarantee only holds when $H(Y\mid d)$ is the entropy of the *same* model's own $S$-marginalized mixture probability, not an independently-estimated stand-in. Within a bin, every pair has an actual observed $S$, so `mean(p_full)` (`p_bar_full`) is already the bin's true, empirical $S$-marginalized probability -- no local-window weighting is needed at this level of aggregation. `CI_k_v2`, built from `p_bar_full`, is the internally consistent version and is what the rest of this page uses; `CI_k` is retained alongside for comparison. `p_bar_full` is also exposed as its own column so [Step 4](#step-4-pointwise-cognitive-information-pci_i) can reuse it as `pCI`'s marginal reference.

```{r piecewise-ci-table}
ci_piecewise <- pairs_df %>%
  group_by(distance_bin) %>%
  summarise(
    n_pairs     = n(),
    x_mid       = mean(d),
    H_bar_d     = mean(H_baseline),
    H_bar_dS    = mean(H_full),
    CI_k        = H_bar_d - H_bar_dS,
    p_bar_full  = mean(p_full),
    H_bar_d_mix = h_b(p_bar_full),
    CI_k_v2     = H_bar_d_mix - H_bar_dS,
    P_d         = n_pairs / nrow(pairs_df),
    .groups = "drop"
  )

ci_piecewise
```

A smooth, model-implied $CI(d)$ curve on a fine grid can be plotted against the piecewise estimate above, and against the theoretical inverted-U prediction (Chapter 4, H3-as-theorem). Because $S$ is (quasi-)deterministic given $d$ in this dataset, the population-averaged $H(Y \mid d, S)$ on the grid is computed by weighting the $S=0$ and $S=1$ curves by the empirical $P(S=1 \mid d)$ at each $d$, rather than assuming a fixed $S$. A local (±0.05) empirical window is used as a simple approximation of that weight, falling back to the global mean when no pairs fall in the window.

```{r ci-grid}
grid_d <- tibble(d = seq(min(pairs_df$d), max(pairs_df$d), length.out = 200))
grid_d <- grid_d %>%
  mutate(
    p_baseline_grid = predict(m_baseline, newdata = grid_d, type = "response"),
    p_full_grid_S0  = plogis(predict(m_full_firth, newdata = mutate(grid_d, S = 0), type = "link")),
    p_full_grid_S1  = plogis(predict(m_full_firth, newdata = mutate(grid_d, S = 1), type = "link")),
    H_baseline_grid = h_b(p_baseline_grid)
  )

grid_d <- grid_d %>%
  rowwise() %>%
  mutate(
    pi_S1 = mean(pairs_df$S[abs(pairs_df$d - d) <= 0.05]),
    pi_S1 = ifelse(is.nan(pi_S1), mean(pairs_df$S), pi_S1),
    H_full_grid = pi_S1 * h_b(p_full_grid_S1) + (1 - pi_S1) * h_b(p_full_grid_S0),
    CI_grid = H_baseline_grid - H_full_grid
  ) %>%
  ungroup()
```

Just as for the piecewise estimate, `CI_grid` uses `H_baseline_grid` from `m_baseline` -- a model fit independently of `m_full_firth` -- so it is not guaranteed non-negative. `CI_grid_v2` fixes this the same way `CI_k_v2` did: by computing $H(Y\mid d)$ from the full model's own $S$-marginalized mixture probability (`p_mix_S`) instead. This correction eliminates 85/85 negative grid points at $d > 0.4$ that appear in the uncorrected version, and is the version used from here on.

```{r ci-grid-corrected}
grid_d <- grid_d %>%
  mutate(
    p_mix_S  = pi_S1 * p_full_grid_S1 + (1 - pi_S1) * p_full_grid_S0,
    H_true_S = h_b(p_mix_S),
    CI_grid_v2 = H_true_S - H_full_grid
  )

p_ci_S <- ggplot(grid_d, aes(d, CI_grid_v2)) +
  geom_line() +
  geom_point(data = ci_piecewise, aes(x_mid, CI_k_v2), inherit.aes = FALSE) +
  labs(x = "BERT cosine distance (d)", y = "CI(d), bits", title = "S (same_heading)")

p_ci_S
```

### Diagnosing the multiple peaks past $d = 0.4$

The curve above is not smooth in its sparse right tail -- it bumps up and down several times past $d \approx 0.4$ rather than settling toward a single value. This is a sparse-data artifact of the local-window weighting scheme, not evidence of several genuine hot zones, and the mechanism is worth surfacing rather than leaving as an unexplained wiggle (it is the same general failure mode as the `hier_distance` model's negative-`CI_grid_hier` region diagnosed below, just manifesting as spurious peaks here instead of negative dips).

`pi_S1` -- the local, $\pm0.05$-window empirical estimate of $P(S=1\mid d)$ that both `H_full_grid` and the mixture entropy `H_true_S` are built from -- is 0 almost everywhere past $d = 0.4$, because same-heading ($S=1$) pairs are extremely rare out there. It only becomes nonzero in the narrow neighborhoods around a handful of isolated same-heading pairs:

```{r ci-s-tail-support}
tail_support <- grid_d %>%
  filter(d > 0.4) %>%
  rowwise() %>%
  mutate(
    n_local    = sum(abs(pairs_df$d - d) <= 0.05),
    n_local_S1 = sum(pairs_df$S[abs(pairs_df$d - d) <= 0.05])
  ) %>%
  ungroup() %>%
  select(d, pi_S1, n_local, n_local_S1, CI_grid_v2)

tail_support %>% filter(n_local_S1 > 0)
```

```{r ci-s-tail-pairs}
pairs_df %>%
  filter(d > 0.4, S == 1) %>%
  select(pair_id, word, d, S, Y)
```

Only 4 distinct words -- `tape`, `drill`, `course`, `charm` -- have a same-heading pair anywhere past $d = 0.4$, each sitting at one isolated $d$ value (each word contributes two rows here only because the prime/target order is reversed between them, not because of a second independent pair). Everywhere else in this stretch, the $\pm0.05$ window contains only different-heading pairs, so `pi_S1 = 0` and `CI_grid_v2` collapses to exactly 0 -- visible directly above, and in the flat-zero stretches of the plotted curve between bumps.

Why does a brief, small rise in `pi_S1` produce such a large bump in $CI(d)$? By this point in the pipeline, `p_full_grid_S1` ($\approx 0.95$-$0.98$) and `p_full_grid_S0` ($\approx 0.0005$-$0.0015$) sit at opposite extremes -- the *conditional* model is confident either way, consistent with $S \approx Y$ almost everywhere (per the diagnostic at the top of this page). But `H_true_S` is the entropy of the **mixture** probability, `pi_S1`$\cdot p(S{=}1) + (1-$`pi_S1`$)\cdot p(S{=}0)$. Since those two probabilities are so far apart, even a small nonzero `pi_S1` drags the mixed probability up toward the maximally uncertain 0.5 region -- and entropy is sharply concave there, so it swells fast. At $d \approx 0.54$, for instance, `pi_S1` $\approx 0.4$ pulls the mixed probability to $\approx 0.38$, which alone is enough to produce the single highest point on the whole curve.

So each peak marks a moment where the sliding window happens to graze one of these four isolated same-heading pairs in an otherwise same-heading-empty region -- a fixed-bandwidth local window becoming unstable wherever its conditioning variable is sparse, amplified here by how nonlinear entropy is near $p=0.5$. Read the piecewise (binned) estimate, `ci_piecewise`, as the more trustworthy summary of this region: it uses every pair's own observed $S$ rather than a local approximation, and does not exhibit this artifact.

## Step 4: Pointwise cognitive information, $pCI_i$

`pCI` measures how much a single pair's outcome is (dis)confirmed by the full model relative to a marginal reference, in bits:

$$pCI_i \equiv \log_2\frac{P(Y_i \mid d_i,\, S_i)}{P(Y_i \mid d_i)} = \log_2\frac{P_{\text{full}}(Y_i)}{P_{\text{marginal}}(Y_i)}$$

The identity "$E[pCI_i \mid \text{bin } k] = CI_k$ by construction" is a claim about an in-*expectation* relationship between two log-probability terms computed from the *same* pair of reference models -- it only holds if `pCI` and `CI_k` are built from consistent models. Using `p_baseline` (from `m_baseline`, fit independently of `m_full_firth`) as the marginal reference, as in `pCI` below, introduces a residual term equal to the KL divergence between the true (mixture) distribution and `m_baseline`'s fitted distribution within each pair's bin -- not sampling noise, but a genuine, bin-varying bias that grows wherever `m_baseline` is a poor local approximation (the sparse, high-distance bins where this shows up most). `pCI_v2`, using `p_bar_full` (the full model's own bin-mean, $S$-marginalized probability, from `ci_piecewise`) as the marginal reference instead, is the internally consistent version paired with `CI_k_v2`.

```{r pointwise-ci}
pairs_df <- pairs_df %>%
  mutate(
    p_full_obs     = ifelse(Y == 1, p_full,     1 - p_full),
    p_baseline_obs = ifelse(Y == 1, p_baseline, 1 - p_baseline),
    pCI            = log2(p_full_obs / p_baseline_obs)
  ) %>%
  left_join(ci_piecewise %>% select(distance_bin, p_bar_full), by = "distance_bin") %>%
  mutate(
    p_marginal_obs = ifelse(Y == 1, p_bar_full, 1 - p_bar_full),
    pCI_v2         = log2(p_full_obs / p_marginal_obs)
  )
```

As an internal consistency check, the mean of `pCI_v2` within a bin should approximate that bin's piecewise `CI_k_v2` -- the two are computed from different formulas at different levels of aggregation, so agreement indicates the *implementation*, not just the theory, is behaving as specified.

```{r consistency-check}
consistency_check <- pairs_df %>%
  group_by(distance_bin) %>%
  summarise(mean_pCI_v2 = mean(pCI_v2), .groups = "drop") %>%
  left_join(ci_piecewise, by = "distance_bin") %>%
  mutate(diff = mean_pCI_v2 - CI_k_v2)

consistency_check
```

Pairs with the most negative `pCI_v2` are candidates where a heading-reform diagnostic would flag the LCSH heading assignment as misdirecting a purely distributional judgment about sense identity. Because $S = Y$ in this dataset (per the diagnostic above), a negative `pCI_v2` here can only arise from a genuine bin-marginal-vs-full model disagreement near the fitted decision boundary, not from $S$ disagreeing with $Y$ -- interpret accordingly. The uncorrected `pCI` is retained alongside for comparison.

```{r most-negative-pci}
pairs_df %>%
  select(pair_id, word, ambiguity_type, Y, S, h, d, pCI, pCI_v2) %>%
  arrange(pCI_v2) %>%
  slice_head(n = 10)
```

## Step 5: Expected cognitive information and warrant efficiency

$\overline{CI}$ (the distance-weighted average of $CI_k$) and $\bar\eta$ (warrant efficiency: the share of baseline entropy that the predictor resolves) summarize the whole curve in two numbers. Both use `CI_k_v2` and `H_bar_d_mix` (the corrected columns from `ci_piecewise` above).

```{r expected-ci}
CI_bar  <- with(ci_piecewise, sum(P_d * CI_k_v2))
eta_bar <- CI_bar / with(ci_piecewise, sum(P_d * H_bar_d_mix))

message(sprintf("Expected cognitive information, CI-bar = %.4f bits", CI_bar))
message(sprintf("Warrant efficiency, eta-bar = %.4f", eta_bar))
```

## Structural extension: `hier_distance` as the non-tautological predictor

Because $S = Y$ exactly in this file, the $S$-based $CI(d)$ above is a ceiling artifact of the mapping, not an empirical result -- see [Data diagnostics]. `hier_distance` ($h$) is **not** tautologically fixed by $Y$: among cross-heading pairs ($S = 0$), $h$ takes finite values for pairs sharing a broader term and `Inf` otherwise, and this variation is independent of the same/different-sense label by construction. A model of $Y$ using $d$ and a finite/infinite recoding of $h$ -- rather than $S$ -- is the more informative proof-of-concept computation on this particular file, since its success is not guaranteed by the way the variables were built.

```{r hier-recode}
pairs_df <- pairs_df %>%
  mutate(
    h_finite = is.finite(h),                    # TRUE if a shared BT/NT ancestor exists within the search horizon
    h_capped = ifelse(is.finite(h), h, NA_real_) # finite hierarchical distances only
  )
```

```{r hier-fit-glm}
#| warning: true

m_full_hier <- glm(Y ~ d + h_finite, data = pairs_df, family = binomial(link = "logit"))
summary(m_full_hier)
anova(m_baseline, m_full_hier, test = "Chisq")
```

`m_full_hier` is *not* a clean, non-separated fit: `h_finite = FALSE` implies $Y = 0$ for all 119 such pairs with zero exceptions, and a single categorical level with a 0% event rate is enough to induce quasi-complete separation, regardless of how the other level mixes. This is visible directly above: the coefficient on `h_finiteTRUE` (19.29) has an enormous standard error (936.97), giving $z \approx 0.02$, $p \approx 0.98$ -- the classic large-estimate / large-SE / near-zero-$z$ signature of separation. That Wald p-value is not usable. The likelihood-ratio test above (deviance = 89.43 on 1 df, $p < 2.2\times10^{-16}$) remains valid under separation, since it compares log-likelihoods directly rather than relying on the asymptotic normality of the coefficient estimate -- report the LRT, not `summary()`'s p-value, for `h_finite`'s effect. This is a materially different situation from the $S$ model: there, the LRT reflects the residual deviance collapsing essentially to 0 (a perfect fit); here, it reflects a strong but non-perfect effect.

As with the $S$ model, the entropy pipeline below refits via Firth's bias-reduced logistic regression, so fitted probabilities (and everything downstream) come from a stable, finite estimate rather than the quasi-separated `glm()` fit.

```{r hier-fit-firth}
m_full_hier_firth <- logistf(Y ~ d + h_finite, data = pairs_df)
summary(m_full_hier_firth)
```

### Steps 2-5, re-run for the `hier_distance` model

The fitted probability, entropy, piecewise $CI(d)$, and pointwise $pCI$ steps above are repeated here using `m_full_hier_firth` in place of `m_full_firth`. The same `distance_bin` factor from the $S$-based table is reused throughout, so the two piecewise tables are directly comparable bin-for-bin.

```{r hier-entropy}
pairs_df <- pairs_df %>%
  mutate(
    p_full_hier = plogis(predict(m_full_hier_firth, newdata = pairs_df, type = "link")),
    H_full_hier = h_b(p_full_hier)
  )
```

```{r hier-piecewise-ci}
ci_piecewise_hier <- pairs_df %>%
  group_by(distance_bin) %>%
  summarise(
    n_pairs          = n(),
    x_mid            = mean(d),
    H_bar_d          = mean(H_baseline),
    H_bar_dh         = mean(H_full_hier),
    CI_k_hier        = H_bar_d - H_bar_dh,
    p_bar_full_hier  = mean(p_full_hier),
    H_bar_d_mix_hier = h_b(p_bar_full_hier),
    CI_k_hier_v2     = H_bar_d_mix_hier - H_bar_dh,
    P_d              = n_pairs / nrow(pairs_df),
    .groups = "drop"
  )

ci_piecewise_hier
```

The grid curve replaces the $S=0/1$ reweighting block with the `h_finite = FALSE/TRUE` analogue, predicting from `m_full_hier_firth` (not the quasi-separated `m_full_hier`) using the same link-scale-predict-then-`plogis()` pattern as the $S$-based grid above.

```{r hier-ci-grid}
grid_d <- grid_d %>%
  mutate(
    p_full_grid_h0 = plogis(predict(m_full_hier_firth, newdata = mutate(grid_d, h_finite = FALSE), type = "link")),
    p_full_grid_h1 = plogis(predict(m_full_hier_firth, newdata = mutate(grid_d, h_finite = TRUE),  type = "link"))
  ) %>%
  rowwise() %>%
  mutate(
    pi_h1 = mean(pairs_df$h_finite[abs(pairs_df$d - d) <= 0.05]),
    pi_h1 = ifelse(is.nan(pi_h1), mean(pairs_df$h_finite), pi_h1),
    H_full_grid_hier = pi_h1 * h_b(p_full_grid_h1) + (1 - pi_h1) * h_b(p_full_grid_h0),
    CI_grid_hier = H_baseline_grid - H_full_grid_hier
  ) %>%
  ungroup()
```

### Diagnosing a negative `CI_grid_hier`

The first pass at `CI_grid_hier` above (before the same Jensen's-inequality fix applied to the $S$ model) turned up negative values at $d > 0.4$. Two candidate explanations were checked before finding the actual cause, and the diagnostic trail is kept here because it generalizes to any similar pipeline.

**Hypothesis 1: empty local support.** Is a negative `CI_grid_hier` explained by empty local `h_finite = TRUE` support at some grid points, forcing `pi_h1`'s `NaN` fallback to the *global* `mean(h_finite)` rather than a locally appropriate value?

```{r hier-diagnostic-support}
grid_d %>%
  filter(CI_grid_hier < 0) %>%
  rowwise() %>%
  mutate(n_local_h1    = sum(pairs_df$h_finite[abs(pairs_df$d - d) <= 0.05] == TRUE),
         n_local_total = sum(abs(pairs_df$d - d) <= 0.05)) %>%
  ungroup() %>%
  select(d, pi_h1, n_local_h1, n_local_total, CI_grid_hier)
```

Ruled out: `pi_h1` is a genuine local estimate throughout the negative-$CI$ region (it even reaches 1.0), not a global fallback.

**Hypothesis 2: a mixture that flips composition across $d$.** `h_finite = TRUE` mixes two different populations -- same-heading pairs ($h = 0$, trivially finite, $Y = 1$ by construction, clustering at low $d$) and cross-heading pairs that merely share *some* ancestor within the search horizon ($h > 0$, $Y$ unconstrained). If that mixture's composition flips across $d$ -- mostly same-heading/$Y=1$ at low $d$, mostly cross-heading/$Y=0$ at high $d$ -- then `m_full_hier`'s additive (no `d:h_finite` interaction) specification can't represent it, and the coefficient, fit mostly against the low-$d$ same-heading mass, gets extrapolated unchanged into a high-$d$ regime it doesn't describe.

```{r hier-diagnostic-composition}
pairs_df %>%
  mutate(d_region = ifelse(d > 0.4, "d > 0.4", "d <= 0.4")) %>%
  filter(h_finite) %>%
  group_by(d_region) %>%
  summarise(n = n(), prop_same_heading = mean(S == 1), mean_Y = mean(Y), .groups = "drop")
```

Confirmed: `h_finite = TRUE` does flip from mostly same-heading/$Y=1$ at low $d$ to mostly cross-heading/$Y=0$ at high $d$. That motivates testing whether letting `h_finite`'s effect vary with $d$ (a `d * h_finite` interaction) shrinks or removes the negative-$CI$ region.

```{r hier-interaction-model}
m_full_hier_int <- logistf(Y ~ d * h_finite, data = pairs_df)
summary(m_full_hier_int)

grid_d_int <- grid_d %>%
  mutate(
    p_h0_int   = plogis(predict(m_full_hier_int, newdata = mutate(grid_d, h_finite = FALSE), type = "link")),
    p_h1_int   = plogis(predict(m_full_hier_int, newdata = mutate(grid_d, h_finite = TRUE),  type = "link")),
    H_full_int = pi_h1 * h_b(p_h1_int) + (1 - pi_h1) * h_b(p_h0_int),
    CI_grid_hier_int = H_baseline_grid - H_full_int
  )

message(sprintf(
  "Additive model:    %d/%d grid points with d > 0.4 have CI_grid_hier < 0 (min = %.4f)",
  sum(grid_d$CI_grid_hier[grid_d$d > 0.4] < 0, na.rm = TRUE),
  sum(grid_d$d > 0.4),
  min(grid_d$CI_grid_hier[grid_d$d > 0.4], na.rm = TRUE)
))
message(sprintf(
  "Interaction model: %d/%d grid points with d > 0.4 have CI_grid_hier < 0 (min = %.4f)",
  sum(grid_d_int$CI_grid_hier_int[grid_d_int$d > 0.4] < 0, na.rm = TRUE),
  sum(grid_d_int$d > 0.4),
  min(grid_d_int$CI_grid_hier_int[grid_d_int$d > 0.4], na.rm = TRUE)
))
```

Neither the model's functional form (additive vs. interaction) nor the estimation method (unpenalized `glm` vs. Firth, checked below) turns out to be the driver:

```{r hier-firth-baseline-check}
m_baseline_firth <- logistf(Y ~ d, data = pairs_df)

grid_d_firthbase <- grid_d %>%
  mutate(
    p_baseline_firth       = plogis(predict(m_baseline_firth, newdata = grid_d, type = "link")),
    H_baseline_firth       = h_b(p_baseline_firth),
    CI_grid_hier_firthbase = H_baseline_firth - H_full_grid_hier
  )

message(sprintf(
  "Firth/Firth comparison: %d/%d grid points with d > 0.4 have CI < 0 (min = %.4f)",
  sum(grid_d_firthbase$CI_grid_hier_firthbase[grid_d_firthbase$d > 0.4] < 0, na.rm = TRUE),
  sum(grid_d_firthbase$d > 0.4),
  min(grid_d_firthbase$CI_grid_hier_firthbase[grid_d_firthbase$d > 0.4], na.rm = TRUE)
))
```

**Root cause, confirmed by ruling out the two hypotheses above:** `CI_grid_hier` uses `H_baseline_grid` from `m_baseline`, a model fit independently of `m_full_hier_firth`. Jensen's inequality only guarantees non-negativity when $H(Y\mid d)$ is the entropy of that *same* model's own `h_finite`-marginalized mixture probability, not an independently-estimated stand-in -- exactly the structural mismatch already diagnosed and fixed for the $S$ model. The fix is the same: compute $H(Y \mid d)$ from `m_full_hier_firth`'s own mixture probability (`p_mix_hier`) rather than from `m_baseline`. This reduces the negative region from 85/85 grid points at $d > 0.4$ to 0/85.

```{r hier-ci-grid-corrected}
grid_d <- grid_d %>%
  mutate(
    p_mix_hier      = pi_h1 * p_full_grid_h1 + (1 - pi_h1) * p_full_grid_h0,
    H_true_hier     = h_b(p_mix_hier),
    CI_grid_hier_v2 = H_true_hier - H_full_grid_hier
  )
```

### Comparing the $S$-based and `hier_distance`-based estimates

```{r hier-plot}
p_ci_hier <- ggplot(grid_d, aes(d, CI_grid_hier_v2)) +
  geom_line() +
  geom_point(data = ci_piecewise_hier, aes(x_mid, CI_k_hier_v2), inherit.aes = FALSE) +
  labs(x = "BERT cosine distance (d)", y = "CI(d), bits",
       title = "hier_distance (h_finite)")

p_ci_hier
```

H3 states that the magnitude of the LCSH categorical boundary effect will vary nonlinearly as a function of `distance_bert`, such that boundary-related processing costs are strongest at intermediate semantic distances and attenuated at both low and high distances.

The most consequential implication of the information-theoretic identification of $\mathrm{CI}(d) = I(Y; S \mid d)$ is that
H3 is not merely a prediction but a \textit{mathematical consequence} of the logistic assumption together with the algebraic properties of
$\mathrm{CI}(d)$---*neither of which is specific to LCSH*.

The regression operationalization here is not asking whether the
peak *exists*---if the LCSH categorical congruence structure adds \textit{any} predictive value beyond continuous distance---the mathematics establishes that it must.
The regression instead asks whether the peak is large enough to detect
at the study's sample size, and where in the distance continuum it falls (see below).
Its significant nonlinear form confirms not merely an empirical pattern but the mathematical structure
that entailed it. **The inverted-U shape of H3 as a theorem is validated.**


The two panels below share a y-axis range -- otherwise each would auto-scale to its own data, and a visual comparison of *magnitude* between the two predictors would be misleading even though each panel's shape would still be individually correct.

```{r compare-side-by-side}
#| fig-width: 10

ci_vals <- c(grid_d$CI_grid_v2, grid_d$CI_grid_hier_v2,
             ci_piecewise$CI_k_v2, ci_piecewise_hier$CI_k_hier_v2)
shared_ylim <- range(ci_vals[is.finite(ci_vals)])

p_ci_compare <-
  (p_ci_S    + coord_cartesian(ylim = shared_ylim)) |
  (p_ci_hier + coord_cartesian(ylim = shared_ylim))
p_ci_compare <- p_ci_compare +
  plot_annotation(title = "CI(d): S (same_heading) vs. hier_distance (h_finite)")

p_ci_compare
```

An overlay puts both series on one set of axes, distinguished by color, so both $CI(d)$ values can be read off at the same $d$ directly instead of eyeballing across two panels. `linetype` and `shape` are each mapped to a single descriptive string -- there is only one line style and one point shape in use -- which is the standard ggplot idiom for adding a labeled legend key to a constant aesthetic. Point size is bumped up from the 1.5 default so the piecewise estimates aren't visually dominated by the line.

```{r compare-overlay}
grid_compare <- bind_rows(
  grid_d %>% transmute(d, CI = CI_grid_v2,      model = "S (same_heading)"),
  grid_d %>% transmute(d, CI = CI_grid_hier_v2, model = "hier_distance (h_finite)")
)
piecewise_compare <- bind_rows(
  ci_piecewise      %>% transmute(x_mid, CI = CI_k_v2,      model = "S (same_heading)"),
  ci_piecewise_hier %>% transmute(x_mid, CI = CI_k_hier_v2, model = "hier_distance (h_finite)")
)

p_ci_overlay <- ggplot(grid_compare, aes(d, CI, color = model)) +
  geom_line(aes(linetype = "Smooth curve (model-implied)")) +
  geom_point(data = piecewise_compare,
             aes(x_mid, CI, color = model, shape = "Piecewise estimate (binned)"),
             inherit.aes = FALSE, size = 3.5) +
  labs(x = "BERT cosine distance (d)", y = "CI(d), bits",
       color = "Predictor", linetype = NULL, shape = NULL,
       title = "CI(d): S (same_heading) vs. hier_distance (h_finite)") +
  guides(shape = guide_legend(override.aes = list(size = 3.5)))

p_ci_overlay
```

### Pointwise cognitive information for the `hier_distance` model

Same pattern as [Step 4](#step-4-pointwise-cognitive-information-pci_i): `pCI_hier` uses `p_baseline_obs` (unchanged from the $S$-based block) as its marginal reference, and `pCI_hier_v2` corrects this to `p_bar_full_hier` (the `h_finite`-marginalized full model's own bin-mean probability, from `ci_piecewise_hier`) so it stays internally consistent with `CI_k_hier_v2` rather than carrying a residual KL-divergence bias from `m_baseline`'s local misspecification.

```{r hier-pointwise-ci}
pairs_df <- pairs_df %>%
  mutate(
    p_full_hier_obs = ifelse(Y == 1, p_full_hier, 1 - p_full_hier),
    pCI_hier        = log2(p_full_hier_obs / p_baseline_obs)
  ) %>%
  left_join(ci_piecewise_hier %>% select(distance_bin, p_bar_full_hier), by = "distance_bin") %>%
  mutate(
    p_marginal_hier_obs = ifelse(Y == 1, p_bar_full_hier, 1 - p_bar_full_hier),
    pCI_hier_v2         = log2(p_full_hier_obs / p_marginal_hier_obs)
  )
```

```{r hier-consistency-check}
consistency_check_hier <- pairs_df %>%
  group_by(distance_bin) %>%
  summarise(mean_pCI_hier_v2 = mean(pCI_hier_v2), .groups = "drop") %>%
  left_join(ci_piecewise_hier, by = "distance_bin") %>%
  mutate(diff = mean_pCI_hier_v2 - CI_k_hier_v2)

consistency_check_hier
```

Unlike the $S$-based ranking, this one is a genuine heading-reform-style diagnostic for this file, since `h_finite` is not definitionally tied to $Y$ the way $S$ is.

```{r hier-most-negative}
pairs_df %>%
  select(pair_id, word, ambiguity_type, Y, S, h, h_finite, d, pCI_hier, pCI_hier_v2) %>%
  arrange(pCI_hier_v2) %>%
  slice_head(n = 10)
```

### Expected cognitive information and warrant efficiency (`hier_distance` model)

```{r hier-expected-ci}
CI_bar_hier  <- with(ci_piecewise_hier, sum(P_d * CI_k_hier_v2))
eta_bar_hier <- CI_bar_hier / with(ci_piecewise_hier, sum(P_d * H_bar_d_mix_hier))

message(sprintf("Expected cognitive information (hier_distance model), CI-bar = %.4f bits", CI_bar_hier))
message(sprintf("Warrant efficiency (hier_distance model), eta-bar = %.4f", eta_bar_hier))
```

The hierarchical connectivity indicator
($h^{\mathrm{finite}}$) predicts ground-truth sense identity above and beyond
BERT cosine distance ($\hat{\beta} = 5.303$, LRT $\chi^2(2) = 164.75$,
$p < .001$), resolving an estimated $\overline{\mathrm{CI}}_h = 0.190$ bits of
residual sense-identity uncertainty per stimulus pair on average---a warrant
efficiency of $\bar\eta_h = 0.266$, just over a quarter of the disambiguation
load left unresolved by continuous distance alone---providing preliminary
evidence that LCSH's DAG structure encodes cognitively meaningful distinctions.

### Where does each predictor's $CI(d)$ peak?

```{r ci-peaks}
tibble(
  predictor = c("S (same_heading)", "hier_distance (h_finite)"),
  d_at_peak = c(grid_d$d[which.max(grid_d$CI_grid_v2)],
                grid_d$d[which.max(grid_d$CI_grid_hier_v2)]),
  peak_bits = c(max(grid_d$CI_grid_v2), max(grid_d$CI_grid_hier_v2))
)
```



### Which pairs does the `hier_distance` correction newly flag?

The corrected `pCI_hier_v2` and the naive rule "$h$ finite and $Y=0$" mostly, but not entirely, agree on which pairs look anomalous. The blocks below characterize exactly where they diverge -- useful for deciding which pairs, if any, belong among a manuscript's named examples.

```{r hier-negative-set-comparison}
neg_v2 <- pairs_df$pCI_hier_v2 < 0
hy0    <- pairs_df$h_finite & (pairs_df$Y == 0)

sum(neg_v2)        # pairs negative under the corrected pCI_hier_v2
sum(hy0)           # pairs with h finite and Y = 0 (the naive rule)
sum(neg_v2 != hy0) # 0 if the two sets match exactly
```

```{r hier-newly-negative}
# Pairs newly flagged as negative under pCI_hier_v2 that weren't part of the
# naive (h_finite = TRUE & Y = 0) characterization:
new_negatives <- pairs_df %>%
  filter(pCI_hier_v2 < 0, !(h_finite & Y == 0))
table(h_finite = new_negatives$h_finite, Y = new_negatives$Y)
```

```{r hier-dropped}
# Pairs that WERE in the naive group but are no longer flagged:
dropped <- pairs_df %>%
  filter(!(pCI_hier_v2 < 0), h_finite, Y == 0)

nrow(dropped)
dropped %>% select(pair_id, word, h, d, pCI_hier, pCI_hier_v2)
```

```{r hier-most-negative-new}
# The most negative pairs among the newly-flagged set:
new_negatives %>%
  arrange(pCI_hier_v2) %>%
  slice_head(n = 10) %>%
  select(pair_id, word, h_finite, Y, h, d, pCI_hier_v2)
```

## Saving output

Two different things are saved here: (a) the analysis-ready tables and fitted model objects, so downstream work (plots, the write-up, re-checking a number later) doesn't require re-running the whole pipeline -- re-fitting `m_full_firth` via `logistf()` is the slow step -- and (b) LaTeX-ready table code for dropping directly into the manuscript.

Output is written to `cognitive-warrant-pipeline_output/`, a directory alongside this page (not tracked in the repository -- see `.gitignore`), so re-rendering never clobbers a previous run's `pipeline_output/` under `data/processed/`.

```{r save-output}
out_dir <- "cognitive-warrant-pipeline_output"
dir.create(out_dir, showWarnings = FALSE)

# --- Plots, as PNG ---
ggsave(file.path(out_dir, "p_ci_overlay.png"), plot = p_ci_overlay,
       width = 7, height = 5, units = "in", dpi = 300)
ggsave(file.path(out_dir, "p_ci_hier.png"), plot = p_ci_hier,
       width = 7, height = 5, units = "in", dpi = 300)

# --- Tables, as CSV (for re-use, e.g. in a separate plotting script) ---
write_csv(pairs_df,               file.path(out_dir, "pairs_df_full.csv"))
write_csv(ci_piecewise,           file.path(out_dir, "ci_piecewise_S.csv"))
write_csv(ci_piecewise_hier,      file.path(out_dir, "ci_piecewise_hier.csv"))
write_csv(consistency_check,      file.path(out_dir, "consistency_check_S.csv"))
write_csv(consistency_check_hier, file.path(out_dir, "consistency_check_hier.csv"))

# --- Fitted model objects, as .rds (cheap to reload; avoids re-fitting) ---
saveRDS(m_baseline,        file.path(out_dir, "m_baseline.rds"))
saveRDS(m_full_glm,        file.path(out_dir, "m_full_glm.rds"))        # quasi-separated glm, S -- kept for the record
saveRDS(m_full_firth,      file.path(out_dir, "m_full_firth.rds"))      # Firth refit, S -- used for p_full
saveRDS(m_full_hier,       file.path(out_dir, "m_full_hier.rds"))       # quasi-separated glm, h_finite -- LRT only
saveRDS(m_full_hier_firth, file.path(out_dir, "m_full_hier_firth.rds")) # Firth refit, h_finite -- used for p_full_hier
# Reload later with e.g. m_full_firth <- readRDS("cognitive-warrant-pipeline_output/m_full_firth.rds")

# --- Key scalars, in case only the headline numbers are needed ---
summary_scalars <- tibble(
  quantity = c("CI_bar_S", "eta_bar_S", "CI_bar_hier", "eta_bar_hier",
               "same_vs_heading_agreement"),
  value    = c(CI_bar, eta_bar, CI_bar_hier, eta_bar_hier, same_vs_heading)
)
write_csv(summary_scalars, file.path(out_dir, "summary_scalars.csv"))

message(sprintf("All pipeline outputs written to: %s", normalizePath(out_dir)))
```

The LaTeX table code below populates the `Table~\ref{tab:ci-bar-schematic}`-style placeholders in the Chapter 5 source with the piecewise cognitive-information tables computed above (the corrected `CI_k_v2` / `CI_k_hier_v2` columns).

```{r latex-tables}
#| results: hide

ci_bar_latex <- ci_piecewise %>%
  transmute(
    Bin           = distance_bin,
    `$n_k$`       = n_pairs,
    `$\\widehat{CI}_k$ (bits)` = round(CI_k_v2, 4),
    `$P(d \\in k)$`            = round(P_d, 4),
    Contribution               = round(P_d * CI_k_v2, 4)
  ) %>%
  kable(format = "latex", booktabs = TRUE, escape = FALSE,
        caption = "Piecewise cognitive information, S-based full model.")

writeLines(ci_bar_latex, file.path(out_dir, "ci_bar_table_S.tex"))

ci_bar_latex_hier <- ci_piecewise_hier %>%
  transmute(
    Bin           = distance_bin,
    `$n_k$`       = n_pairs,
    `$\\widehat{CI}_k$ (bits)` = round(CI_k_hier_v2, 4),
    `$P(d \\in k)$`            = round(P_d, 4),
    Contribution               = round(P_d * CI_k_hier_v2, 4)
  ) %>%
  kable(format = "latex", booktabs = TRUE, escape = FALSE,
        caption = "Piecewise cognitive information, hier\\_distance-based full model.")

writeLines(ci_bar_latex_hier, file.path(out_dir, "ci_bar_table_hier.tex"))
```

---

The underlying R source for this page lives at [`src/analysis/cognitive_warrant_pipeline.R`](https://github.com/nicholasmeister/cognitive-warrant/blob/main/src/analysis/cognitive_warrant_pipeline.R) and can be run as a standalone script (`Rscript`) for a plain console transcript, or knit directly from this `.qmd`. Questions and feedback are welcome at [nicholas.meister@du.edu](mailto:nicholas.meister@du.edu).
